---
title: "Alex on Tokens.x"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
# install.packages("plotly")
library(plotly)

knitr::opts_chunk$set(echo = TRUE)
```

## Alex on Tokens.x

This is the analysis of the tracing output obtained from a run of Alex on its `Tokens.x` example.

### Closure evaluation

Note that the dataset contains calls from many modules, not just Alex itself.

```{r load, echo = FALSE}
data  <- read_tsv( '/mnt/analysis/home/viluon/alex-dyn-trace-valid.tsv' #'/mnt/analysis/paks/alex-dyn-trace-valid.tsv'
                 , lazy = FALSE
                 # 30 c's at the end for the argument columns
                 , col_types = "iiccciiciicccccccccccccccccccccccccccccc"
                 #                  ||||||
                 #                  ||||++-- source & destination addresses (for update frames)
                 #                  ||||
                 #                  |||+---- result diff
                 #                  ||+----- arity
                 #                  |+------ closure lifetime
                 #                  +------- closure type
                 )
diffs <- select(filter(data, type == "argument diff"), !c("src address", "dst address", type, "thread id"))

# helper
parseFunId <- function (tibl) {
  tibl %>%
    mutate(
      package = gsub("^(.+)_[A-Z].*$",                         "\\1", `function`, perl = TRUE),
      module  = gsub("^.+_([A-Z]\\w*(?:\\.[A-Z]\\w*)*).*$",    "\\1", `function`, perl = TRUE),
      binder  = gsub("^.+_[A-Z]\\w*(?:\\.[A-Z]\\w*)*\\.(.*)$", "\\1", `function`, perl = TRUE)
    )
}
```

```{r initial exploration, echo = FALSE}
unique_call_signs <- diffs %>%
  #select(!timestamp) %>%
  #distinct(`function`, arg1, arg2, arg3, arg4, arg5, arg6, arg7, arg8, arg9, arg10, arg11, arg12, arg13)
  group_by(`function`, arity, arg1, arg2, arg3, arg4, arg5, arg6, arg7, arg8, arg9, arg10, arg11, arg12, arg13) %>%
  summarise(calls = n()) %>%
  group_by(`function`)
```

How many function calls preserve laziness?
```{r laziness-preserving calls}
funcs_only <- unique_call_signs %>% filter(arity > 0)
arg_cols   <- paste0("arg", 1:12)
all_calls  <- sum(funcs_only$calls)

# *ack* I gotta check how does if_all() behave with NAs :(
lazy <- funcs_only %>%
  filter(!if_any(all_of(arg_cols), \(x) grepl("->", x)) & !if_all(all_of(arg_cols), \(x) grepl("(con|<not pointery>)", x)))
lazy_calls <- sum(lazy$calls)

unboxed <- funcs_only %>%
  filter(if_any(all_of(arg_cols), \(x) grepl("<not pointery>", x)))
unboxed_calls <- sum(unboxed$calls)

p <- \(n) prettyNum(n, big.mark = " ")
report <- \(msg, n, m) show(paste0(round(100 * n / m, 2), "% ", msg, " (", p(n), " / ", p(m), ")"))
report("of all calls are somewhat lazy", lazy_calls, all_calls)
report("of all calls involve unboxed values or void", unboxed_calls, all_calls)
```

What does closure evaluation produce?

```{r result shapes}
unique_result_shapes <- diffs %>%
  # let's look at functions only
  filter(arity > 0) %>%
  mutate(`result atoms` = gsub("(\\w+)Atom \\d+", "\\1Atom", `result atoms`)) %>%
  group_by(`result atoms`) %>%
  summarise(n = n()) %>%
  mutate(`result arity` = str_count(`result atoms`, ",") + 1) %>%
  arrange(., -n)

single_result_calls <- unique_result_shapes %>% filter(`result arity` == 1) %>% summarise(sum(n))
report("of all calls produce a single value", single_result_calls, all_calls)

shape_count <- \(shape) unique_result_shapes %>%
    filter(grepl(shape, `result atoms`)) %>%
    summarise(sum(n))

report("of all calls end in a pointer to a constructor", shape_count("con"), all_calls)
report("of all calls end in a pointer to a thunk", shape_count("thunk"), all_calls)
```

How long do closures live?

```{r closure lifetimes}
entries <- select(filter(data, type == "closure entry"), c(timestamp, "function", "closure type", lifetime))

entries %>%
  filter(`closure type` == "thunk") %>%
  ggplot(., aes(x = lifetime)) +
  geom_histogram() +
  ggtitle("thunk lifetimes") +
  xlim(0, 2000000) +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_y_continuous(labels = scales::label_comma())

entries %>%
  filter(`closure type` != "thunk") %>%
  ggplot(., aes(x = lifetime)) +
  ggtitle("non-thunk lifetimes") +
  geom_histogram() +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_y_continuous(labels = scales::label_comma())

hist = entries %>%
  ggplot(aes(x = lifetime + 1, fill = `closure type`)) +
  ggtitle("all lifetimes") +
  geom_histogram() +
  scale_x_continuous(labels = scales::label_comma(), trans = "log2") +
  scale_y_continuous(labels = scales::label_comma())

ggplotly(hist)

tlifetime_counts =
  entries %>%
  filter(`closure type` == "thunk") %>%
  group_by(lifetime) %>%
  summarise(n = n())

with_sum = tlifetime_counts %>%
  mutate( sum = sum(tlifetime_counts$n)
        , p = 100 * n / sum
        ) %>%
  arrange(-n)

X = 4

thunks_evaluated_within_X_steps = with_sum %>%
  ungroup() %>%
  filter(lifetime < X) %>%
  summarise(n = sum(n))

with_sum
thunks_evaluated_within_X_steps

report(paste("of thunks are evaluated within", X, "closure entries from the point of their creation"), thunks_evaluated_within_X_steps$n[1], with_sum$sum[1])
```

There are non-thunk closures with lifetimes close to 0. What are those?
```{r short-lived closures}
short_lived = select(filter(data, type == "closure entry"), c(timestamp, "function", "closure type", lifetime)) %>%
  filter(lifetime < 8)

short_lived %>%
  ggplot(., aes(fill = `closure type`, x = `lifetime`)) +
  ggtitle("Short-lived heap objects") +
  geom_bar(position = "dodge") +
  # scale_x_continuous(labels = scales::label_comma()) +
  scale_y_continuous(labels = scales::label_comma())

all_heap_objects = (data %>%
  filter(type == "closure entry") %>%
  summarise(n = n()))$n[1]

young_paps_thunks = short_lived %>%
  filter(lifetime < 3, `closure type` %in% c("thunk", "pap")) %>%
  summarise(n = n())

report("of all heap objects are paps and thunks with lifetimes < 3", young_paps_thunks$n[1], all_heap_objects)
```



TODO:

- [ ] find always-strict functions
- [ ] find sometimes-strict functions
- [ ] find a way to link that information back to the source code
- [ ] which functions are primarily responsible for triggering evaluations?
- [ ] which modules?

```{r consistently strict functions}
# consistently strict in the first argument: either triggers evaluation
# or receives a constructor in the first place
strictInArg1 <- funcs_only %>%
  filter(any(grepl("(con|->)", arg1)))
```

```{r inconsistently strict functions}
# inconsistently strict in the first argument
########################################
# FIXME: this ignores funs and paps,   #
#        which is specific to this     #
#        dataset                       #
########################################
is_thunk  <- \(c) any(c == "thunk")
is_strict <- \(c) any(grepl("->", c))
inconsistently_strict <- funcs_only %>%
  filter(is_thunk (c_across(starts_with("arg")))) %>%
  filter(is_strict(c_across(starts_with("arg"))))
report("of all calls are to inconsistently strict functions", sum(inconsistently_strict$calls), all_calls)


inconsistently_strict %>%
  parseFunId() %>%
  ungroup() %>%
  group_by(package, module, binder) %>%
  summarise(n = sum(calls)) %>%
  arrange(., -n)

# the top inconsistently strict function appears to be IntSet intersection
focus <- function(m, n) {
  c <- funcs_only %>%
    ungroup() %>%
    parseFunId() %>%
    filter(grepl(m, module) & grepl(n, binder)) %>%
    select(arity)
  ar <- c[[1]][1]

  funcs_only %>%
    ungroup() %>%
    parseFunId() %>%
    filter(grepl(m, module, perl = TRUE) & grepl(n, binder, perl = TRUE)) %>%
    select(!starts_with("arg"), num_range("arg", 1:ar)) %>%
    select(!c(package, `function`, arity)) %>%
    arrange(., -calls)
}

focus("", "unpackA")
```


```{r modules responsible for evaluation}
inconsistently_strict %>%
  parseFunId() %>%
  ungroup() %>%
  select(!`function`) %>%
  group_by(package, module) %>%
  filter(if_any(all_of(arg_cols), \(x) grepl("->", x))) %>%
  # FIXME: imprecise: if a call evaluates multiple arguments, only one is counted
  summarise(evaluations = sum(calls)) %>%
  arrange(., -evaluations)
```


```{r rarely lazy functions}
# sum calls of the same laziness signature
# and find functions which have more strict calls than lazy calls
inconsistently_strict %>%
  pivot_longer(starts_with("arg"), names_to = "position", names_pattern = "arg(\\d*)", values_to = "diff") %>%
  filter(!is.na(diff)) %>%
  group_by(`function`, arity, position, diff) %>%
  summarise(n = sum(calls)) %>%
  ungroup() %>%
  parseFunId() %>%
  select(!`function`) %>%
  group_by(package, module, binder) %>%
  # at the following step, the number of calls to a function is equivalent
  # to the sum of the newly added diff columns in any of the function's rows
  pivot_wider(names_from = diff, values_from = n) %>%
  filter(thunk < `thunk->con`) %>%
  ungroup() %>%
  arrange(., -`thunk->con`)

focus("", "wgo2_ssMk")
```









































